---
title: "ODE Examples.Rmd"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Enforced oscillations

$$
  z'' = -g/L\ \sin(z) + \frac{\omega^2}{L}\ Y\cos(z)\sin(\omega\ t)
$$

```{r}
fun <- function(t, z) {
    g <- 9.81
    L <- 1.0
    Y <- 0.25
    w <- 2.5
    dz1 <- z[2]
    dz2 <- -g/L*sin(z[1]) + w^2/L*Y*cos(z[1])*sin(w*t)
    c(dz1, dz2)
}

sol <- pracma::ode45(fun, 0, 60, c(0, 0))

plot(sol$t, sol$y[, 1], type = 'l', lwd = 1.5, col = "darkblue",
     ylim = c(-1, 1), xlab = "time", ylab = "elongation",
     main = "Force acting on a pendulum")
grid()
```

------------------------------------------------------------------------

## Lotka-Volterra equation

$$
  \begin{align}
  dx/dt &= a\ (1 - \frac{x(t)}{x_{max}})\ x(t) - b\ x(t)\ y(t)\\
  dy/dt &= -c\ y(t) + d\ x(t)\ y(t)
  \end{align}
$$

```{r}
fn1 <- function(t, y, p) {
    a <- p[1]; b <- p[2]
    c <- p[3]; d <- p[4]
    dx <- a*(1 - y[1]/3.0)*y[1] - b*y[1]*y[2]
    # dx <-  a*y[1] - b*y[1]*y[2]
    dy <- -c*y[2] + d*y[2]*y[1]
    c(dx, dy)
}

fn2 <- function(t, y, p) {
    a <- p[1]; b <- p[2]
    c <- p[3]; d <- p[4]
    # dx <- a*(1 - y[1]/2.0)*y[1] - b*y[1]*y[2]
    dx <-  a*y[1] - b*y[1]*y[2]
    dy <- -c*y[2] + d*y[2]*y[1]
    c(dx, dy)
}
p0 <- c(2/3, 4/3, 1, 1)
ic <- c(1.0, 1.0)

sol1 <- pracma::ode45(fn1, 0, 50, ic, p = p0)
sol2 <- pracma::ode45(fn2, 0, 50, ic, p = p0)

plot(c(0, 50), c(0, 2.0), type = 'n',
     xlab = "time", ylab = "", main = "Lotka-Volterra equations")
lines(sol1$t, sol1$y[, 1], lty = 2, col = 1)
lines(sol1$t, sol1$y[, 2], lty = 2, col = 2)

lines(sol2$t, sol2$y[, 1], lwd = 2, col = 1)
lines(sol2$t, sol2$y[, 2], lwd = 2, col = 2)

grid()
```

------------------------------------------------------------------------

## Van-der-Pol equation

$$
  y'' - \mu (1 - y^2)\ y' + y = 0
$$

```{r}
vdP <- function(t, y, mu) {
    dy1 <- y[2]
    dy2 <- mu*(1-y[1]^2)*y[2] - y[1]
    c(dy1, dy2)
}

mu1 <- 0.4; y1 <- c(0.1, 0.0)
mu2 <- 2.0; y2 <- c(1.0, 1.0)
sol1 = pracma::ode45(vdP, 0, 30, y1, mu = mu1)
sol2 = pracma::ode45(vdP, 0, 30, y2, mu = mu2)

plot(c(0, 30), c(-3, 3), type = 'n',
     xlab = "time", ylab = "", main = "van-der-Pol equation")
lines(sol1$t, sol1$y[, 1], lwd = 1.5, col = "navy")
lines(sol2$t, sol2$y[, 1], lwd = 1.5, col = "firebrick")
grid()
```

------------------------------------------------------------------------

## Trajectory of a soccer ball

$$
  \begin{align}
  x''(t) &= -r\ v\ x'(t)\\
  y''(t) &= -r\ v\ y'(t) - g
  \end{align}
$$

When the second derivatives are eliminated, we gat the following system
of differential equations:

$$
  \begin{align}
  y_1' &= y_2\\
  y_2' &= -r\ v\ y_2\\
  y_3' &= y_4\\
  y_4' &= -r\ v\ y_4 - g
  \end{align}
$$

where $v = v(t) = \sqrt{x'^2 + y'^2} = \sqrt{y_2^2 + y_4^2}$ is the
speed of the ball. We assume a drag coefficient of $cw = 0.09$ and a
cross-sectional area of $A = 0.0365$ [m\^2] for a diameter of $d = 21.5$
cm for a standardized soccer ball.

```{r}
# Berechne Luftwiderstand
cw = 0.4
rho = 1.2
A = 0.0365
( r = cw * rho/2 * A )
```

```{r}
g <- 9.81
sball <- function(t, y) {
    r <- 0.009
    v <- sqrt(y[2]^2+y[4]^2)
    dy1 <- y[2]
    dy2 <- -r*v*y[2]
    dy3 <- y[4]
    dy4 <- -r*v*y[4] - g
    c(dy1, dy2, dy3, dy4)
}
```

```{r}
y0 <- c(0, 20, 0, 20)
sol <- pracma::ode45(sball, 0, 5, y0, atol = 1e-8)

traj <- function(t) pracma::interp1(sol$t, sol$y[, 3], t)
t1 <- uniroot(traj, c(1, 5))$root

sol <- pracma::ode45(sball, 0, t1, y0, atol = 1e-8)

plot(sol$y[, 1], sol$y[, 3], type = 'l', col = "darkblue")
grid()
```

```{r}
w = 45  # angle in degrees

y0 <- c(0, 28*pracma::cosd(w), 0, 28*pracma::sind(w))
sol <- pracma::ode45(sball, 0, 5, y0, atol = 1e-8)

traj <- function(t) pracma::interp1(sol$t, sol$y[, 3], t)
t1 <- uniroot(traj, c(1, 5))$root

sol <- pracma::ode45(sball, 0, t1, y0, atol = 1e-8)

plot(sol$y[, 1], sol$y[, 3], type = 'l', col = "darkblue")
grid()
```

\# 35 51.69 m \# 36 52.10 m \# 38 52.57 m \# 39 52.81 m \# 40 52.88 m \#
41 52.94 m \# 42 53.00 m \# 43 52.95 m \# 44 52.84 m \# 45 52.71 m \# 46
52.55 m \# 48 51.98 m

------------------------------------------------------------------------

## The 'diffeqr' package

We have to assume that Julia is installed on the computer and available
through the PATH environment variable.

### The flame model again

```{r}
# library(diffeqr)
de <- diffeqr::diffeq_setup()

f <- function(u, p, t) u^2 - u^3

u0 <- 1/250
tspan <- c(0, 500)

probl <- de$ODEProblem(f, u0, tspan)
sol <- de$solve(probl)
```

```{r}
plot(sol$t, sol$u, type = 'l', col = "darkred")
points(sol$t, sol$u, pch=20, col = "darkgray")
grid()
```

NOTE: The 'diffeqr' package has been archived on CRAN, but I guess it
will resurrect within a short time period.

### Chemical reaction

Another example: There are three chemical substances in a reactor with
initial concentrations of 0.6, 0.2, and 0.2. The chemical engineer tells
us that the concentrations of the substances change in the presence of
the other substances according to the following laws:

$$
  \begin{align}
  c_1' &= -k1c1 + k3c3\\
  c_2' &= +k1c1 - k2c2\\
  c_3' &= +k2c2 - k3c3
  \end{align}
$$

The inital concentrations are $k1=0.3, k2=0.2, k3=0.5$.The
concentrations after a suitable reaction time are to be determined.

```{r}
# library(diffeqr)
de <- diffeqr::diffeq_setup()

conc <- function(u, p, t) {
  du1 <- -p[1]*u[1] + p[3]*u[3]
  du2 <-  p[1]*u[1] - p[2]*u[2]
  du3 <-  p[2]*u[2] - p[3]*u[3]
  c(du1, du2, du3)
}
```

```{r}
u0 <- c(0.3,  0.2,  0.5)   # initial concentrations
p0 <- c(0.55, 0.15, 0.30)  # reaction constants
tspan <- c(0, 20)          # 20 minutes

problem  <- de$ODEProblem(conc, u0, tspan, p0)
solution <- de$solve(problem)
```

It can be convenient to turn this solution structure into an R matrix
with sapply, having each row as a time series. The transpose of the
matrix can be turned into a data frame.

```{r}
mat <- sapply(solution$u, identity)

udf <- as.data.frame(t(mat))
```

```{r}
matplot(solution$t, udf, type = 'l', col = c(1, 2, 4),
        lwd = 1.5, lty = 1,
        xlab = "time", ylab = "concentration", main = "Reaction profile")
grid()
```

### The SIR model for spread of disease

Epidemics are often model through so-called SIR models with

-   $S = S(t)$ \# no. of susceptible persons
-   $I = I(t)$ \#no. of infected persons
-   $R = R(t)$ \#no. of recovered persons

Here we treat one of the simplest SIR models, the Kermack-McKendrick
model as a system of differential equations.

$$
  \begin{align}
  S'(t) &= - \alpha\ I(t)\ S(t)\\
  I'(t) &= \alpha\ I(t)\ S(t) - \beta\ I(t)\\
  R'(t) &= \beta\ I(t)
  \end{align}
$$

The corresponding R function is

```{r}
SIRmod <- function(t, y, p) {
  a <- p[1]; b <- p[2]
  dy1 <- -a*y[2]*y[1]
  dy2 <-  a*y[2]*y[1] - b*y[2]
  dy3 <-  b*y[2]
  c(dy1, dy2, dy3)
}
```

and for instance we solve it with the following parameters and initial
conditions.

```{r}
a <- 0.6; b = 0.35
y0 <- c(0.99999, 0.00001, 0)
tspan <- c(0, 100)

sol <- pracma::ode45(SIRmod, 0, 100, y0, p = c(a, b), atol = 1e-10)

plot(c(0, 100), c(0, 1), type = 'n',
     xlab = "days", ylab = "number of ... persons", main = "SIR model")
grid()
lines(sol$t, sol$y[, 1], col = 1, lwd = 1.5)
lines(sol$t, sol$y[, 2], col = 2, lwd = 1.5)
lines(sol$t, sol$y[, 3], col = 3, lwd = 1.5)
legend(0, 0.8
       , legend = c("susceptible", "infected", "recovered"),
       , col = c(1,2,3), lwd = 1.5, bty = 'n')
```

Find more realistic parameters by solving the system of equations
numerically, for a variety of initial conditions.

See the 'odin' package for more advanced SIR models (deterministic and
stochastic ones) and plots of their solutions.

------------------------------------------------------------------------

## Boundary Value Problems (BVP)

### Deflection of a beam

$$
  y'' - \frac{T}{E\ I}\ y = \frac{wx(x-L)}{2\ E\ I}
$$

```{r}
L <- 100
w <- 100
E <- 10e07
T <- 500
I <- 500
```

```{r}
beam <- function(x, y) {
  dy1 <- y[2]
  dy2 <- T/(E*I)*y[1] + w*x*(x-L)/(2*E*I)
  c(dy1, dy2)
}
```

```{r}
sol <- pracma::rk4sys(beam, 0, L, c(0, 0), n = 100)
```

Solving BVP with `pracma::bvp`

```{r}
f <- function(x) rep(0, length(x))
g <- function(x) rep(T/(E*I), length(x))
h <- function(x) w*x*(x-L)/(2*E*I)

sol <- pracma::bvp(f, g, h, c(0, L), c(0, 0))

plot(sol$xs, -sol$ys, type = 'l', col = "darkblue", lwd = 2)
```

Solve with package 'bvpSolve'

```{r}
beam2 <- function(x, y, parms) {
  dy1 <- y[2]
  dy2 <- T/(E*I)*y[1] + w*x*(x-L)/(2*E*I)
  list(c(dy1, dy2))
}

init <- c(0, NA)
end  <- c(0, NA)

sol <- bvpSolve::bvptwp(yini = init, x = seq(0, L, length.out = 101),
                fun = beam2, yend = end)

xs <- sol[, 1]
ys <- sol[, 2]
plot(xs, -ys, type = 'l')
```

### Droplet

The height of the surface of a water droplet satisfies

$$
  y'' = (y-1)*(1+y'^2)^{1.5}
$$

with $y(1.0) = y(-1.0) = 0$. We try to solve it with `bvpsolve` from the
'bvpSolve' package.

```{r}
droplet <- function(t, y, params) {
  dy1 <- y[2]
  dy2 <- (y[1] - 1) * (1 + y[2]^2)^(3/2)
  return( list(c(dy1, dy2)) )
}

init <- c(0, NA)
end  <- c(0, NA)

sol <- bvpSolve::bvpcol(yini = init, x = seq(-1, 1, length.out = 101),
                          fun = droplet, yend = end)

xs <- sol[, 1]
ys <- sol[, 2]
plot(xs, ys, type = 'l', col = 4, asp = 1, bty = 'n',
     xlab = "", ylab = "", main = "Form of a droplet")
grid()
polygon(c(xs, -1), c(ys, 0), col = 4, border = "gray50")
```
